
# Introduction: Fundamentals of R 

SOLUTION

Rmarkdown files combine chunks of R code with Markdown syntax. 'Knitting' the file runs the code blocks and renders the Markdown syntax into a nicely formatted output document (.pdf, .docx, or .html).

Some Markdown basics:

*Italics* **Bold**

* unordered list items
1. ordered list item

Wrapping text in '$\$' tells Markdown to use LaTeX Equation Syntax.

$y = b_{0} + b_{1} x$

Don't worry too much about this, you'll mostly be answering by filling in code blocks or writing out text answers.

Code Blocks: 

You can insert code blocks (in a variety of languages but we'll stick to R here). They look like this:
```{r}
# This is a comment

#declaring variables 
#These both do the same thing, the arrow is preferred style in R but either are acceptable
x <- 5 
x = 5
```

```{r}
# can print with 'print()' or just calling variable name
print(x)
x
```

**Vectors** - all elements must be of the same type
```{r}

vec = c(1,2,3,4,5)
vec[3]
length(vec)
vec = append(vec,10) #adding values
vec
vec = c(vec,20) #another way to add values
vec

#Can create sequences of numbers with seq(), or rep()
vals = seq(1:10)
vals

zeros = rep(0,10)
zeros

y = rep(c(1,2), times= 4)
y
```

To get extra information about a function just put '?' in front and information will pop up in the help window
```{r}
?seq()
```

**Lists** - can contain elements of different types and values can be named
```{r}
lst = list("emily", 10, TRUE, "Sam", 6)
print(lst) #not the prettiest output

lst_named = list(name = "Jerry", major = "CS", year = 2023)
lst_named$major
```

Can concatenate lists together
```{r}
lst1 = list("apple", "banana", "pear")
lst2 = list("carrot", "celery")
concat = c(lst1, lst2)
concat
```


**For Loops and If Statements:**
```{r}
newSeq = seq(1:9)
for(val in newSeq){
  if(val%%2==0){
    print(paste( val, "is even")) #must use paste() to print strings and variables in same line
  }
  else{
   print(paste(val, "is odd"))
  }
}
```


### Problem 1: Put all the multiples of 15 from 1 to 1000 in a vector (save it as a variable named multiples_15)
```{r}
multiples_15 = seq(from=15, to =1000, by =15)
multiples_15
```

```{r}
multiples_15 = c()
for(i in  1:1000){ #inclusive on both bounds
  if(i%%15==0){
    multiples_15 = append(multiples_15,i)
  }
}
multiples_15
```

**Data Frames** - list of vectors of equal length, most common way of storing data in R. We normally won't be creating data frames from scratch but here's some code in case it's necessary. 
```{r}
#Data Frames
names = c("Emily", "Jerry", "Sam", "Kevin")
years = c(2022,2023,2022,2023)
isCSmajor = c(FALSE,TRUE,FALSE,FALSE)
tas = data.frame(col1 = names, col2 = years, col3 = isCSmajor)
tas
```
```{r}
#changing column names
colnames(tas) = c("Name", "Year", "Is CS Major")
tas
```

We will mostly be working with data frames constructed from reading csv (comma separated value) files. More to come next week!


Some useful functions:

```{r}
v = seq(1:20)
mean(v)
min(v)
max(v)
median(v)
sd(v)
sum(v)
```


**Matrices:**

R has a lot of functionality to help deal with matrices. We list some below
```{r}
mat = matrix(c(1,2,3,4),2,2) #vector of values, number of rows, number of columns
mat
nrow(mat)
ncol(mat)
mat_byrow = matrix(c(1,2,3,4),2,2,byrow=TRUE) # fills values row wise
mat_byrow
mat_byrow[1,] #first row, all the columns
mat_byrow[,1] #first column, all the rows

```
```{r}
rbind(c(1,2,3), c(4,5,6)) 
cbind(c(1,2,3),c(4,5,6))
```

Generating random numbers:


```{r}
# generate 10 random numbers between 0 and 10
rand = sample(0:25,10,replace=TRUE) #range (inclusive on both ends), number of values needed, replace?
rand
```



## Problem 2: Create a 20x10 matrix where each row is a random integer array with a minimum of 1, max of 9 (inclusive). Then for each row, sum all the values and add the sum to a vector of values called 'random_sums'

```{r}
rand = sample(1:9,200,replace=T)
mat = matrix(c(rand),20,10,byrow=T)

random_sums = c()
for (i in 1:nrow(mat)){
  random_sums = append(random_sums,sum(mat[i,]))
}
random_sums
```

```{r}
mat = matrix(c(sample(0:9,200,replace=TRUE)),20,10)
#mat
random_sums = c()
for (row in 1:nrow(mat)){
  random_sums = append(random_sums, sum(mat[row,]))
}

random_sums
```



**Set Operations:**

Can preform various set operations on two vectors in R
```{r}
vec1 = c(10,20,30,40)
vec2 = c(10,15,20,25,30)
union(vec1,vec2)
intersect(vec1,vec2)
setdiff(vec1,vec2)
```

## Problem 3: Find the numbers that are in both 'multiples_15' and 'random_sums' and put them in an array called both

```{r}
intersect(multiples_15,random_sums)
```


**Final Note**: We know that you might not have any experience in Python, R, or Computer Science/Math in general, and that's completely fine. Don't be intimidated - this is a part of the process! Please come to office hours, and one of the instructors would give you a crash course. Additionally, we plan to host a Python workshop soon so that you can bridge the gap, if any. If you still feel underprepared, please come talk to one of the instructors - we will work it out.
