
# Data Visualization


## NETID: FILL IN HERE

Now that we have a basic understanding of R and data manipulation, it is time to learn how to represent data visually!

This lecture is comprised of the following components:

* Importance of Data Visualization
* Introduction to `ggplot2`
* Basic visualization techniques
* Visualization Tips and Tricks
* Advanced visualization techniques


## Importance of Data Visualization

As has been widely established, it is much easier for the human brain to comprehend and process information when it is in the form of graphs, charts, etc. than when it is presented in esoteric spreadsheets and verbose reports. Therefore, when working with large amounts of data, it is crucial to step back and make sense of the data - and data visualization helps do just that!

More specifically, Data Visualization is important for the following reasons:

* Informative: A lot can be told in just a small space!
* Appealing: A visualization is much more visually appealing than loads of data
* Universal: It helps in universal communication!
* Predictive: It is much easier to observe patterns and predict certain trends via a visualization


Note that there are typically two types of visualizations: *distributional* (using histograms or box plots to assess the distribution of a variable) and *correlational* (using line plots or scatter plots to understand the relationship between two variables).

There are some plotting capabilities built right into R but they are generally limited. To get nicer plots we can use a library called `ggplot2`. It's in the `tidyverse` library (remember that from last week?).

```{r}
library(tidyverse)
```

Here's a helpful cheatsheet on ggplot2: https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf


How `ggplot2` works is that it builds the graph in layers. You start with a call to the `ggplot()` function itself, pass it the data frame we are working with, and the mapping of variables to visual marks with `aes()` which stands for aesthetics. 


Before we jump into ggplot2, let's take a look at the built in plotting capabilities of R. These will work just fine, they just aren't as pretty or as fun as `ggplot2`

```{r}
#first we will use the built in iris data set

library(datasets)
data(iris)
summary(iris)
```

### Bar Graph


A bar graph is one of the most common techniques and is widely used to convey information in an easy to understand manner! 

It is used to display data that has been classified into nominal or ordinal categories. It represents the respective frequencies/magnitudes for these categories in a comparative manner. Typically, one axis has numeric values (which may be the frequencies in most cases) and the other axis has the categories.

The bar graph for the dataset has been plotted below, using the built in barplot function:

```{r}
bar1 = mean(iris$Sepal.Length)
bar2 = mean(iris$Sepal.Width)
bars = c(bar1,bar2)
barplot(bars,  # vector of counts to plot
        main = "Average of Sepal Features Bar Graph", # main title
        xlab = "Feature", # x-axis label
        ylab = "Average (cm)", # y-axis label
        names.arg = c("Sepal Length", "Sepal Width"), # category names for the x-axis
        col = "darkgreen", # color of the bars
        ylim = range(0,6)) # resetting the y-axis limit (try removing this and investigating why this was added)

```

Great! We have a bar graph that compares the average length and width of sepals in the given dataset.

### Histogram

A histogram is very similar to a bar graph - but it represents a *continuous* variable on its x-axis rather than discrete units like a bar graph. Because it is continuous, we need to specify how many bars we need in total, and in order to do so, we need to set the number of bars to display, called breaks. Now, the built in histogram function in R does something a little funky with the breaks argument. If you give it a single number to indicate the number of bars, it takes that as a suggestion and gives 'pretty' values for the breaks which may result in a different number of bars than what was passed to breaks. Sometimes this is fine and other times we might want direct control over the number of bars, or their exact break points. You can pass break a vector of the exact points to create bars at but this requires a little preprocessing of the data to determine the best values for this vector. 


## Problem 1

Enter the code to make a histogram with a title and axis labels and 10 breaks. The X axis would be have the 'Sepal Width' column, and the Y axis would have the frequency.

(hint: https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/hist)

```{r}
# FILL IN HERE
```


We can also built a histogram using `ggplot2`. Here we have access to a parameter called "binwidth" that sets the width of each of the bins (a bit easier to work with than breaks from above)


```{r}
 #call our normal ggplot function, add on the geom_histogram 
ggplot(iris, aes(x=Sepal.Width))+ geom_histogram(binwidth = 0.1,fill="#69b3a2", color="#e9ecef", alpha=0.9) + ggtitle("Sepal Width")
```


While we will use a random number (i.e. 0.1) as our bin width for ease of demonstration, choosing the correct bin size can be very important at times. It also has some interesting statistical reasoning behind it! (http://www.statisticshowto.com/choose-bin-sizes-statistics/)


### Density Plots

While histograms are popular, density plots are also favored for several reasons

* Histogram shape varies wildly depending on the bin width
* Density plots smooth out outliers and local fluctuations

The second point can be a weakness however, since local fluctuations can be very important. Let's look at an example (we'll be using `ggplot2` more now!):

```{r}
 #call our normal ggplot function, add on the geom_density
ggplot(iris, aes(x=Sepal.Width))+ geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.9)+ggtitle("Sepal Width")
```

The smoothing parameter for the density plot can be adjusted by changing something called the smoothing bandwidth (`adjust`) in order to adjust for sensitivity to fluctuations.

```{r}
ggplot(iris, aes(x=Sepal.Width))+ geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.9, adjust = 0.3)+ggtitle("Sepal Width")
```
```{r}
ggplot(iris, aes(x=Sepal.Width))+ geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.9, adjust = 2)+ggtitle("Sepal Width")
```

As you can see, the bandwidth value can have a drastic effect on the resulting graph. These are extreme examples, however; appropriate values for bandwidth differ case by case.

### Box Plot


A box plot shows the max, min, and the median values of one feature from a given dataset. It is a great tool when a quick summarization is necessary or when the spread of data needs to be shown. For example, it can be used to visualize test scores.

An important point to note is that an *outlier* is a data point that is 1.5 times above the upper quartile or 1.5 times below the lower quartile.

```{r}
ggplot(iris, aes(x=Species, y = Sepal.Length)) + geom_boxplot()+ggtitle("Sepal Length by Species")
```

### Violin Plot

A violin plot is very similar to a box plot. The only big difference is that while box plot has a box (which represents IQR: Inter Quartile Range, a range between 25% percentile and 75% percentile of the values) and two whiskers to represent maximum and minimum, violin plot has a three lines that indicates max (100% percentile), median (50% percentile), and min (0% percentile) and also has a *probability density distribution* of the data at different values along the line.

```{r}
ggplot(iris, aes(x=Species, y = Sepal.Length, fill = Species)) + geom_violin()+ggtitle("Sepal Length by Species")
```


```{r}
#can nest the box plot within the violin plot by just adding the geom_boxplot function and making it a bit smaller so it fits inside
ggplot(iris, aes(x=Species, y = Sepal.Length, fill = Species)) + geom_violin()+geom_boxplot(width=0.1)+ggtitle("Sepal Length by Species")
```
### Scatter Plot

Scatter plots can be very useful for visualizing information for two or more features (that correspond to the number of dimensions). They are very useful for extrapolating information.

```{r}
ggplot(iris, aes(x=Sepal.Length, y = Sepal.Width, col = Species)) + geom_point()+ggtitle("Sepal Width vs. Sepal Length colored by Species")
```


It is noticeable that

* There is a clear clustering behavior for setosa (pink)
* Versicolor and virginica are not clearly separated (blue/green)

This kind of information is valuable in assessing what kind of model to choose, and what kind of additional analysis that needs to take before we are sure of what to do with our dataset. It is possible, however, to further embellish our visual analysis by using a 3d plot and adding in another feature. The 3d scatterplot in R uses a new package, `plotly`. This library creates a more interactive, somewhat more complicated visualization.

```{r}
# If you want to knit your notebook to a word doc, you have to comment this part out. Word doesn't know how to handle displaying this kind of plot

library(plotly)
plot_ly(iris, x = ~Sepal.Length, y = ~Sepal.Width, z = ~Petal.Length, color = ~Species)
```

Looking at this 3d scatterplot, we can see that the versicolor and virginica is actually much more separable, unlike what was indicated in our previous 2d plot. We can therefore conclude that these three features are enough to implement an effective classifier.


### Heat Map

Another plot for visualizing density is a heat map. Heat maps are a bivariate distribution which assigns colors to different regions depending on the density (or frequency or magnitude, depending on which feature you are trying to visualize) of values in that region.

```{r}
ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width) ) +
  geom_bin2d(bins = 10) +
  scale_fill_continuous(type = "viridis") +
  theme_bw()
```

(not quite as pretty as the python version, sorry :( )

### Correlation Plot

Similar to a heat map is a correlation plot. However, instead of looking two variables to find their corrleation, we examine all the variables to determine which are correlated. To create this plot, we need to subset our data set to get only the continuous variables and convert it to a matrix. 


```{r}
data = iris[,c(1,2,3,4)]
head(data)
cormat <- round(cor(data),2)
head(cormat)
```


```{r}
library(reshape2)
melted_cormat <- melt(cormat) #converts the data into one column so we can use it
#head(melted_cormat)

ggplot(data = melted_cormat, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile() + 
geom_text(aes(Var2, Var1, label = value), color = "white", size = 4)
```
## Problem 2

Download the temperatures dataset from the course website and make sure it is in the same folder as this file. Now pick one of the above graph types to visualize one or more columns from the temperatures dataset. Hint: the `unique` function may come in handy to reduce the number of points.


```{r}
# FILL IN HERE
```


## Visualization Tips and Tricks

### Changing Color

Let's take the bar graph we programmed earlier and change its color!

It's fairly straightforward to do so: Use the attribute `col` when you're creating the barplot, and specify the color of choice! Here's an example that turns the bar plot pink

```{r}
bar1 = mean(iris$Sepal.Length)
bar2 = mean(iris$Sepal.Width)
bars = c(bar1,bar2)
barplot(bars, main = "Average of Sepal Features Bar Graph", xlab = "Feature", ylab = "Average (cm)", names.arg = c("Sepal Length", "Sepal Width"), col = "thistle", ylim = range(0,6))
```

If you run `colors()` in the R console, it will print out all of the color keywords that are built into R. You can also pass hexidecimal values to the color argument to get custom colors. 

We can change the theme of colors in a ggplot scatter plot

```{r}
ggplot(iris, aes(x=Sepal.Length, y = Sepal.Width, col = Species)) + geom_point()+ggtitle("Sepal Width vs. Sepal Length colored by Species")+ scale_color_manual(values = c("#00AFBB", "#E7B800", "#FC4E07"))
```

The `Rcolorbrewer` package also gives many predefined, pretty color scales you can access after reading in the package (try `display.brewer.all()` in the console). Let's apply it to our box plots from earlier

```{r}
library(RColorBrewer)
ggplot(iris, aes(x=Species, y = Sepal.Length, fill = Species)) + geom_boxplot()+ggtitle("Sepal Length by Species") +scale_fill_brewer(palette = "BuPu") 
```

`ggplot2` comes in very handy in that it automatically adds the legend for us (something done manually in python). This is another reason to lean towards using the package rather than the built in R functions when plotting. We can adjust the position of these legends pretty easily

```{r}
ggplot(iris, aes(x=Species, y = Sepal.Length, fill = Species)) + geom_boxplot()+ggtitle("Sepal Length by Species") +scale_fill_brewer(palette = "BuPu")+theme(legend.position = "top")
```


## Subplots

```{r}
# Creating dummy data
x = seq(0, 3*pi, length.out = 500)
y = sin(x^1.5)
df = data.frame(x,y)
```

2 subplots
```{r}
require(gridExtra) #need this package for the arrange function
line = ggplot(data=df, aes(x=x, y=y, group =1)) +
  geom_line()
scatter = ggplot(data=df, aes(x=x,y=y)) + geom_point()
grid.arrange(line, scatter, ncol=2) #need to arrange the plots on a grid to display them side by side

```

4 subplots

```{r}
line = ggplot(data=df, aes(x=x, y=y, group =1)) +
  geom_line()
scatter = ggplot(data=df, aes(x=x,y=y)) + geom_point(size = 0.5)
dens = ggplot(data=df, aes(x=y)) + geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.9, adjust = 0.5)
scatter2 = ggplot(data=df^2, aes(x=x,y=y)) + geom_point(size = 0.5)
grid.arrange(line, scatter, dens, scatter2,ncol=2) #need to arrange the plots on a grid to display them side by side
```
## Problem 3

Tweak your previous graph (or create a new one) so it includes one of the above tips and tricks.

```{r}
#FILL IN HERE
```


Note: You are by no means required to memorize all (or any for that matter) of these functions. Visualizing data isn't always the most straightforward process in R and this notebook should serve as a good reference guide. I've also found this website to be very helpful: https://www.r-graph-gallery.com/index.html

As always, please come to office hours or catch me before/after lecture with any questions!


