
# Introduction to Classifiers


NETID: FILL IN HERE


Two lectures ago we covered linear regression and predicting the value of a continuous variable. We use *classifiers* to predict binary or categorical variables. Classifiers can help us answer yes/no questions or categorize an observation into one of several categories. 

## kNN Classifier

There are various classification algorithms, each of which is better suited to some situations than others. In this lecture we are learning about *kNN*, which is one of these classifiers

```{r}
df = read.csv('lecture6data.csv')
df = subset(df, select = -c(X,id) )
head(df)
```

## Problem 1

Build a kNN model predicting whether an observation is benign or malignant. You should split the dataset into a training set and a test set as covered previously in the course, fit the model on the observations in the training set, and predict the target variable for the test set.

There are a couple of things to note for this problem. First, you are free to choose whichever features you want to predict the target feature, but you should not use id or the target variable itself. Second, you can optionally choose the k parameter for the kNN model (the default value is 5).

Save your predictions in a variable named "predictions".

**Please do not change the variable names already provided as they are used later in the demo**

```{r}
library(class) #you might need to download this package if you haven't before
train = sample() #FILL IN HERE
train_df = #FILL IN HERE, drop the target column (hint: df[..., -1])
test_df = #FILL IN HERE, drop the target column (hint: df[..., -1])

train_df_label = #FILL IN HERE - we want to extract out JUST the target column (hint: df[...,1])
test_df_label = #FILL IN HERE - we want to extract out JUST the target column (hint: df[...,1])

classifier_knn = knn() #FILL IN HERE - use k = 10, after you've loaded in the class package you can type '??knn' into the R console to see the specifications of the knn function and determine the arguments, basically you'll need to pass in the train_df, test_df, the train labels and your value for k.

confusion_matrix = table(test_df_label, classifier_knn) #I'll fill this in for you but be sure to look at the output!
confusion_matrix
```


```{r}
misclassification_error = mean(classifier_knn != test_df_label) #percentage of points we misclassified
accuracy = 1-misclassification_error #accuracy is 1-percentage of points we misclassified
accuracy
```


## Measuring Accuracy

Measuring the accuracy of classifiers is more intuitive than calculating the accuracy of a linear regression model. When we predict categorical values, our accuracy score is simply the proportion of values that we computed correctly. For example, if we have a test set of size 100 and we predict 93 of the observations correctly, we have an accuracy score of 93 percent.

```{r}
print(paste("accuracy score from our model: ", accuracy))

#compute the accuracy of predicting all observations as Benign
baseline = rep('B',114)
misclassification_error_baseline = mean(classifier_knn != baseline) #percentage of points we misclassified
accuracy_baseline = 1-misclassification_error_baseline #accuracy is 1-percentage of points we misclassified
print(paste("accuracy score from predicting Benign for every entry: ", accuracy_baseline))

improvement = (accuracy - accuracy_baseline)/accuracy_baseline #See our improvement 
print(paste("Improvement: ", improvement))
```

The above improvement shows just how beneficial the kNN model can be. It also shows us that we have chosen an appropriate value for k because there is an improvement over the baseline assumption (average category of values).


## Fit/Overfitting

Below are accuracy scores of the same kNN model, but with the value of k changing. Note how the accuracy changes as k increases. As mentioned during the lecture, a high value of k can improve the accuracy of the model, but too high a value of k will essentially be the average of all of the data (if this block isn't running, be sure you didn't change any of the variable names above).

```{r}
set.seed(42)
#k = 1
classifier_knn_1 = knn(train_df, test_df, cl=train_df_label, k = 1)
misclassification_error_1 = mean(classifier_knn_1 != test_df_label)
accuracy_1 = 1-misclassification_error_1

#k = 10
classifier_knn_10 = knn(train_df, test_df, cl=train_df_label, k = 10)
misclassification_error_10 = mean(classifier_knn_10 != test_df_label)
accuracy_10 = 1-misclassification_error_10

#k = 50
classifier_knn_50 = knn(train_df, test_df, cl=train_df_label, k = 50)
misclassification_error_50 = mean(classifier_knn_50 != test_df_label)
accuracy_50 = 1-misclassification_error_50

print(paste("Accuracy Score when k = 1: ", accuracy_1))
print(paste("Accuracy Score when k = 10: ", accuracy_10))
print(paste("Accuracy Score when k = 50: ", accuracy_50))

```

## Problem 2

Now we are going to plot the relationship between the value of k and the accuracy score of the model for this data set.

Using a loop, create models with neighbors ranging from 1-30. Find the accuracy for each of these models and graph them with number of neighbors on the x-axis and accuracy on the y-axis. *You do not need to redo the train test split.*

```{r}
accuracies = c()

for (i in 1:30){
 #FILL IN HERE
}
```

```{r}
plot()#FILL IN HERE
```



## Confusion Matrix


            | Positive       | Negative
            | (predicted)    | (predicted)
  ---------------------------|---------------
   Positive | True Positive  | False Negative
   (Actual) |                |                   
  ----------|----------------|---------------
   Negative | False Positive | True Negative
   (Actual) |                |    
   
   
   
Here are the equations specified in the lecture for your convenience with the next problem.

**Sensitivity** = True Positive /(True Positive + False Negative)

**Specificity** = True Negative /(True Negative + False Positive)

**Accuracy** = (True Positive + True Negative) / Total

**Error** = (False Positive + False Negative) / Total

**Precision** = True Positive / (True Positive + False Positive) 


## Problem 3

Given the Table Below, Calculate the **Specificity**, **Sensitivity**, **Overall Error Rate**, **Overall Accuracy**, **Precision** of the data.
            
            
            | Positive       | Negative
            | (predicted)    | (predicted)
  ---------------------------|---------------
   Positive |     146        |      32
   (Actual) |                |                   
  ----------|----------------|---------------
   Negative |     21         |      590
   (Actual) |                |    
   
   
```{r}
#FILL IN HERE
```
             
            
## Problem 4 (extra credit)

### Part a
Before running kNN, which of the following kinds of preprocessing should we do? Choose all that apply.

1) Scale

2) Center

3) Remove correlated features

4) Remove outliers


### Part b
Let's say you want to run kNN on a dataset with both continuous features and binary features. Can you think of any potential issues that might arise from mixing these data types? Also, how might you preprocess categorical data to use in a kNN? Ordinal data?


## Problem 5 (extra credit)

We've talked about sensitivity and specificity. Recall these high level intuitions:
- high sensitivity -> able to correctly identify positives
- high specificty -> able to correctly identify negatives

### Part a
Identify a model that has have 100% sensitivity, no matter what dataset it is run on. Similarly, identify a model with 100% specificity.

**Hint**: Recall that a "model" is just a function, meaning that it takes in an input and *spits out an output*. Your job is to figure out, if you get an input x, should the output for that x be 0 or should it be 1?

### Part b
In Problem 2, you plotted kNN accuracy vs `k`. Now, make a plot kNN sensitivity vs. `k` and another plot for kNN specificity. Use the same dataset as in Problem 2, and go from k=1 to k=30.

### Part c
Now, plot the average of specificity and sensitivity against the number of neighbors.
            
### Part d (just for fun)
Prove the following statement:

*If the testing set has the same number of positive and negative examples, then the accuracy is equal to the average of the specificity and sensitivity.*


## Problem 6 (0 points)

In INFO1998, we focus on the high level concepts and applications, and as a result, we rarely delve into the computations that go into our machine learning algorithms. This isn't to say that those computations are unimportant. When working with *enormous* datasets, some algorithms become *infeasible* since they need to do so much *computation* that training a model could take several days. When working with **complex** datasets with **special properties**, we can sometimes **adapt** algorithms based on our understanding of the algorithm's *computations*.

This question challenges you to think about the computations involved in kNN.

Also, this is worth 0 points, so feel free to look up the answer if you want.

### Part a
When a kNN model is making a prediction for a sample, what does it need to do? Be specific.

### Part b
In a past class, saw that the linear regression model is just a linear function; that is, the whole model can be represented by just a couple numbers (the weights/coefficients). Based on Part a, what data is necessary to represent a trained kNN model?

### Part c
Based on Part b, describe the training algorithm for a kNN model. Recall that a training algorithm is how you go from a training set to a representation of a model. Hint: it's super simple.

In parts D and E, write your answers in terms of the quantities: 
- number of training samples, T
- number of samples in dataset, N
- number of features in dataset, F
- number of neighbors, k

### Part d
Write down an expression estimating the memory needed to represent a trained kNN model.

### Part e
Write down an expression estimating the time needed for kNN to make a prediction for a single point.

### Part f
Estimate the memory space needed to represent a kNN with the specifications below. Also estimate how long it would take to predict 1,000 test samples.

- number of training samples: 1,000,000
- number of samples in dataset: 100,000,000
- number of features in dataset: 50
- number of neighbors: 5
- size of one feature of one sample: 8 bytes
- time to calculate distance for x features: x/10,000 seconds

### Part g
There are many variations on kNN that aim to speed up kNN predictions. This might involve saving less of the train set, checking only a subset of the saved data, encoding data differently, or organizing data differently. Look up two of these variations, and compare their advantages/disadvantages. 

